---
title: "Konza_Vegetation_Analysis"
author: "Jamison Foglesong"
format: html
execute:
  echo: false
  Warning: false
  error: false
---

## Introduction

This analysis will look at vegetation data from the Konza National Ecological Observatory Network (Neon) site located near Manhattan, KS. It will use tidyverse packages to wrangle and visualize some of the data.

## Loading the data

The data are provided in csv files. We'll use the tidyverse package `readr` to load it into R.

```{r}
#| label: read-csv

# load packages

library("readr")
library("dplyr")
library("ggplot2")
library("ggthemes")
library("knitr")

#read in the data
konza_data <- read.csv("data/raw data/NEON.D06.KONZ.DP1.10058.001.div_1m2Data.2025-08.basic.20251222T234053Z.csv")
```

## Data summarizing with `dplyr`

We'll next use the dplyr package to summarize some different aspects of the data.

```{r}
#| label: summarize-with-dplyr

# first summary to make will be a count of the number of observations in each family
konza_data %>%
  group_by(family) %>%
  tally() %>%
  arrange(desc(n)) %>%
  kable()
```

This analysis showed that Poaceae and Asteraceae are among the most common family found in these plots.

```{r}
#| label: mean-pcnt-cover-by-family

#the goal is to calculate the mean percent cover of each family
konza_data %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
    arrange(desc(mean_pcnt_cover)) %>%
  kable()
```

This analysis showed that the largest percent cover is the `NA` taxa, which means it likely wasn't able to be identified, but the highest mean percent cover at %16 across the plots was Fagaceae

```{r}
#| label: filter-by-elevation

#the goal is to calculate the mean percent cover of each family at elevation greater than 390
konza_data %>%
  filter(elevation > 390) %>%
  group_by(family) %>%
  summarise(mean_pcnt_cover = mean(percentCover),
            sd_pcnt_cover = sd(percentCover)) %>%
    arrange(desc(mean_pcnt_cover))
```

Next we'll learn to use `mutate()`, which is a function to create a new column based on contents from an existing column

```{r}
#| label: use-mutate

#select pulls out columns
#mutate creates a new column and can use current column data

konza_data %>%
  select(scientificName, family, percentCover) %>%
  mutate(propCover = percentCover / 100)
```

Next up, we'll use `ggplot2` to make some figures

```{r}
#|label: elevation-percent-by-family

konza_data %>%
  filter(!is.na(family)) %>%
  ggplot(aes(x = percentCover)) +
  geom_histogram() +
  labs(title = "Histogram of percent cover, all species",
       x = "Percent Cover") 
  show.legend = TRUE 
  theme_get()
```


This concludes the sample exercises completed in class.

## Practicing exploratory data analysis

I will be creating 3 plots analyzing the Konza NEON data to practice using `dplyer`
and `ggplot2` to explore and analyze data

## Figure 0 - Diversity by subplot (deprecated)
This figure explores the relative species variation of the subplots observed in the NEON data.

```{r}
#|label: diversity-by-subplot
taxontibble <- tibble(x = konza_data$taxonID)
subplotIDtibble <- tibble(x = konza_data$subplotID)

konza_data %>%  
  group_by(subplotID) %>%
  ggplot(aes(x = subplotIDtibble, y = count(taxontibble, name = "Unique Species"))) +
        scale_x_discrete(TRUE) +
         geom_bar() +
           labs(title = "Count of Unique Species per Subplot", x = "Subplot (ID number)", y = "amount of unique species")
         
```
I am abandoning this figure, I have wasted too much time hyper fixated on how to
format discrete data into something usable for the x-axis of my plot.

## Figure 1 -
```{r}
#|llabel: Land Cover Variables
  
konza_data %>%
  filter(!is.na(otherVariables)) %>%
  ggplot(aes(x = otherVariables)) +
  geom_bar(
    na.rm = FALSE,
           position = position_dodge2(preserve = "single")) +
           theme(
             axis.text.x.bottom = element_text(
               margin = margin(t = 50, unit = "pt"),
              angle = 80))
```


